---
title: "Fitting Exercise"
---

**Data Preparation**

```{r setup, warning=FALSE, message=FALSE}
#load modules
library(here) #for relative paths
library(tidyverse) #for data wrangling
library(ggplot2) #for plots
library(corrplot) #for correlation plot
library(tidymodels) #for model fitting

#raw file URL
url <- "https://raw.githubusercontent.com/metrumresearchgroup/BayesPBPK-tutorial/master/data/Mavoglurant_A2121_nmpk.csv"

#download file
download.file(url,
              destfile = "fitting-exercise/Mavoglurant_A2121_nmpk.csv",
              mode = "wb")

#data path
data_path <- here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv")

#load data
data <- read.csv(here(data_path))

#explore the data
str(data)
head(data, 10)

#write code to make a plot that shows a line for each individual, with DV on the y-axis and time on the x-axis. Stratify by dose (e.g., use a different color for each dose, or facets).
ggplot(data, aes(x=TIME, y=DV, group=ID))+
  geom_line(alpha=0.6, color="darkgrey")+
  geom_point(size=1, color="black")+
  theme_bw()+
  facet_wrap(~ DOSE)

#Write code that keeps only observations with OCC = 1.
data2 <- data %>%
  filter(OCC==1) #only keep OCC=1

#Write code to exclude the observations with TIME = 0, then compute the sum of the DV variable for each individual using dplyr::summarize(). Call this variable Y. The result from this step should be a data frame/tibble of size 120 x 2, one column for the ID one for the variable Y. Next, create a data frame that contains only the observations where TIME == 0. This should be a tibble of size 120 x 17. Finally, use the appropriate join function to combine those two data frames, to get a data frame of size 120 x 18.

data3 <- data2 %>%
  filter(TIME!=0) #excluding TIME=0

sumDV <- data3 %>%
  group_by(ID) %>%
  summarize(Y = sum(DV), .groups = "drop") #DV sum of each individual

time0 <- data2 %>%
  filter(TIME==0) #only observations with time=0

data4 <- left_join(time0, sumDV, by = "ID") #join sumDV and time0

data5 <- data4 %>%
  mutate(
    RACE = factor(RACE),
    SEX  = factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT) #convert SEX and RACE to factors; keep certain variables
```

**Additional EDA**


```{r}
summary(data5)

data5 %>%
  group_by(SEX) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y)
  ) #summary table by SEX

#categorical predictor: SEX
ggplot(data5, aes(x = SEX, y = Y)) +
  geom_boxplot() +
  theme_bw()

#Mean Y is slightly higher on average in SEX=1 compared to SEX=2, but variability is similar and SEX=2 is a drastically smaller sample size

data5 %>%
  group_by(RACE) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y)
  ) #summary table by RACE

#categorical predictor: RACE
ggplot(data5, aes(x = RACE, y = Y)) +
  geom_boxplot() +
  theme_bw()

#Mean Y varies across RACE, but comparisons are difficult with RACE=7 and RACE=88 having drastically smaller sample sizes than RACE=1 and RACE=2

#distribution of Y
ggplot(data5, aes(x = Y)) +
  geom_histogram(bins = 30) +
  theme_bw()
#Y appears slightly skewed. Should check based on DOSE

#distribution of Y (based on DOSE)
ggplot(data5, aes(x = Y, fill = as.factor(DOSE))) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 20) +
  theme_bw() +
  labs(fill = "Dose")
#Some overlap but histograms show Y increases with DOSE

#categorical predictor: DOSE
ggplot(data5, aes(x = as.factor(DOSE), y = Y)) +
  geom_boxplot() +
  theme_bw()
#boxplots support Y increasing by DOSE

#distribution of HT
ggplot(data5, aes(x = HT)) +
  geom_histogram(bins = 30) +
  theme_bw()
#HT relatively normal

#distribution of WT
ggplot(data5, aes(x = WT)) +
  geom_histogram(bins = 30) +
  theme_bw()
#WT relatively normal

#plot AGE VS Y
  ggplot(data5, aes(x = AGE, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
#Little to no relationship between AGE and Y

#plot WT VS Y
ggplot(data5, aes(x = WT, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
#Some negative association between WT and Y

#plot HT VS Y
ggplot(data5, aes(x = HT, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
#Some negative association between HT and Y

#correlation matrix
cor <- data5 %>%
  select(Y, DOSE, AGE, WT, HT) %>%
  cor()
cor

#correlation plot
corrplot(cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)

#Y is strongly positively correlated with DOSE, weakly negatively with HT and WT. HT and WT are moderately correlated with each other.
```

**Model Fitting**

```{r}
set.seed(123) #for reproducibility
lm_mod <- linear_reg()

#Fit a linear model to Y using DOSE
lm_fit1 <- lm_mod %>%
  fit(Y ~ DOSE, data = data5)
lm_fit1 
#Y roughly increases by 58 units for each 1 unit of DOSE

#Fit a linear model to Y using all predictors
lm_fit2 <- lm_mod %>%
  fit(Y ~ ., data = data5)
lm_fit2

#Compute RMSE and R-squared for both models
preds1 <- predict(lm_fit1, new_data = data5) %>%
  bind_cols(data5 %>% select(Y))
rmse(preds1, truth = Y, estimate = .pred)
rsq(preds1, truth = Y, estimate = .pred)
#For the first model (Y ~ DOSE), RMSE is 666 and rsq is 0.52

preds2 <- predict(lm_fit2, new_data = data5) %>%
  bind_cols(data5 %>% select(Y))
rmse(preds2, truth = Y, estimate = .pred)
rsq(preds2, truth = Y, estimate = .pred)
#For the second model (Y ~ .), RMSE is 591 and rsq is 0.62

log_mod <- logistic_reg() %>%
  set_engine("glm")

#Fit a log model to Y using SEX
log_fit1 <- log_mod %>%
  fit(SEX ~ DOSE, data = data5)
log_fit1

#Fit a log model to Y using all predictors
log_fit2 <- log_mod %>%
  fit(SEX ~ DOSE + AGE + RACE + WT + HT, data = data5)
log_fit2

#Compute accuracy and ROC-AUC for both models
pred_class1 <- predict(log_fit1, new_data = data5, type = "class")
pred_prob1  <- predict(log_fit1, new_data = data5, type = "prob")
results1 <- bind_cols(data5 %>% select(SEX), pred_class1, pred_prob1)
accuracy(results1, truth = SEX, estimate = .pred_class)
roc_auc(results1, truth = SEX, .pred_2)
#For the first log model (SEX ~ DOSE), accuracy is 0.867 and ROC AUC is 0.41

pred_class2 <- predict(log_fit2, new_data = data5, type = "class")
pred_prob2  <- predict(log_fit2, new_data = data5, type = "prob")
results2 <- bind_cols(data5 %>% select(SEX), pred_class2, pred_prob2)
accuracy(results2, truth = SEX, estimate = .pred_class)
roc_auc(results2, truth = SEX, .pred_2)
#For the second log model (SEX ~ .), accuracy is 0.94 and ROC AUC is 0.025
```

**Module 10: Model Improvement**

```{r}
##Data Prep
#remove the RACE variable
data6 <- subset(data5, select= -RACE)

#set seed
rngseed = 1234
set.seed(rngseed)

#data split
data_split <- initial_split(data6, prop = 3/4)

#create data frames for the two sets:
train_data <- training(data_split)
test_data <- testing(data_split)

##Model Fitting
#Fit a linear model to Y using DOSE
lm_fit_10.1 <- lm_mod %>%
  fit(Y ~ DOSE, data = train_data)
lm_fit_10.1 
#Y roughly increases by 58 units for each 1 unit of DOSE

#Fit a linear model to Y using all predictors
lm_fit_10.2 <- lm_mod %>%
  fit(Y ~ ., data = train_data)
lm_fit_10.2

##Model Performance Assessment 1
#Compute RMSE and R-squared for both models
preds_10.1 <- predict(lm_fit_10.1, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))
rmse(preds_10.1, truth = Y, estimate = .pred)
rsq(preds_10.1, truth = Y, estimate = .pred)
#For the first model (Y ~ DOSE), RMSE is 703 and rsq is 0.451

preds_10.2 <- predict(lm_fit_10.2, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))
rmse(preds_10.2, truth = Y, estimate = .pred)
rsq(preds_10.2, truth = Y, estimate = .pred)
#For the second model (Y ~ .), RMSE is 627 and rsq is 0.562

#define null model
null_mod <- null_model(mode="regression") %>%
  set_engine("parsnip")
#fit the null model
null_fit <- null_mod %>%
  fit(Y ~ 1, data = train_data)
#compute RMSE
null_preds <- predict(null_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))
rmse(null_preds, truth = Y, estimate = .pred)
#for the null model, RMSE is 948

##Model Performance Assessment 2
#create 10-fold cross-validation splits
folds <- vfold_cv(train_data, v=10)

#perform cross-validation for Model 1 (Y ~ DOSE)
cv_fit1 <- lm_mod %>%
  fit_resamples(
    Y ~ DOSE,
    resamples = folds,
    metrics = metric_set(rmse)
  )
#RMSE 691

#perform cross-validation for Model 2 (Y ~ all predictors)
cv_fit2 <- lm_mod %>%
  fit_resamples(
    Y ~ .,
    resamples = folds,
    metrics = metric_set(rmse)
  )
#RMSE 640

collect_metrics(cv_fit1)
collect_metrics(cv_fit2)

#Now with a new seed
rngseed2 = 5678
set.seed(rngseed2)
#create 10-fold cross-validation splits
folds <- vfold_cv(train_data, v=10)

#perform cross-validation for Model 1 (Y ~ DOSE)
cv_fit3 <- lm_mod %>%
  fit_resamples(
    Y ~ DOSE,
    resamples = folds,
    metrics = metric_set(rmse)
  )
#RMSE 693

#perform cross-validation for Model 2 (Y ~ all predictors)
cv_fit4 <- lm_mod %>%
  fit_resamples(
    Y ~ .,
    resamples = folds,
    metrics = metric_set(rmse)
  )
#RMSE 649

collect_metrics(cv_fit3)
collect_metrics(cv_fit4)
```

# This section added by Talia C. Swanson

## Observed vs Predicted for all 3 

```{r}
# Predictions from the original models
pred_null <- predict(null_fit, new_data = train_data)$.pred
pred_dose <- predict(lm_fit_10.1, new_data = train_data)$.pred
pred_all  <- predict(lm_fit_10.2, new_data = train_data)$.pred

# Combine into long format for plotting
pred_df <- tibble(
  ID = train_data$ID,
  Observed = train_data$Y,
  Null = pred_null,
  Dose = pred_dose,
  All = pred_all
) %>%
  pivot_longer(cols = c("Null", "Dose", "All"),
    names_to = "Model",
    values_to = "Predicted")

# Plot of Observed vs Predicted
ggplot(pred_df, aes(x = Observed, y = Predicted, color = Model, shape = Model)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0,5000), ylim = c(0,5000)) +
  theme_bw() +
  labs(title = "Observed vs Predicted Values",
    x = "Observed Y",
    y = "Predicted Y")
```

The observed vs predicted plot shows how close the model predictions are to the actual values of Y. Ideally, the points should fall along the dashed 45-degree line, which represents perfect predictions. The null model predicts nearly the same value for all observations and does not match the observed data well. The dose-only model performs slightly better but still shows noticeable differences from the observed values. The model with all predictors follows the reference line more closely, suggesting it provides the best predictions among the three models.

## Residual Plot for the All-Predictors Model
```{r}
# Residuals for all-predictors model
resid_all <- train_data$Y - pred_all
resid_df <- tibble(
  Observed = train_data$Y,
  Residuals = resid_all
)

ggplot(resid_df, aes(x = Observed, y = Residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-max(abs(resid_all)), max(abs(resid_all)))) +
  theme_bw() +
  labs(title = "Residuals vs Observed Values (All Predictors Model)",
    x = "Observed Y",
    y = "Residuals")
```

The residual plot shows the difference between the observed and predicted values. Ideally, the residuals should be randomly scattered around zero. In this plot, the residuals are mostly centered around zero, but they appear to spread out more as the observed values increase. This suggests that the model predicts smaller values better than larger values, and some variation in the data is still not explained by the model.

## Bootstrap Prediction and Uncertainty 
```{r}
# reset seed for reproducibility 
set.seed(rngseed) 

# To create 100 bootstrap resamples
boot_splits <- bootstraps(train_data, times = 100)

# Fit model on each bootstrap and predict on the original training data
pred_bs <- map(1:100, function(i){
  bs_data <- analysis(boot_splits$splits[[i]])
  bs_fit <- lm_mod %>% fit(Y ~ ., data = bs_data)
  predict(bs_fit, new_data = train_data)$.pred
}) %>% do.call(cbind, .)

# Compute median and 89% CI for predictions
preds_ci <- apply (pred_bs, 1, quantile, probs = c(0.055, 0.5, 0.945)) |> t() |> as_tibble()
names(preds_ci) <- c("Lower", "Median", "Upper")

# Combine with original predicted values
pred_plot_df <- tibble(
  Observed = train_data$Y,
  Original = pred_all
) %>%
  bind_cols(preds_ci)

# Plot predictions with bootstrap uncertainty
ggplot(pred_plot_df, aes(x = Observed)) +
  geom_point(aes(y = Original), color = "black", size = 2, alpha = 0.6) +
  geom_point(aes(y = Median), color = "blue", size = 2, alpha = 0.6) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.3, color = "blue", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0,5000), ylim = c(0,5000)) +
  theme_bw() +
  labs(title = "Model 2 Predictions with Bootstrap Uncertainty",
    x = "Observed Y",
    y = "Predicted Y")
```

This plot shows the predictions from the full model along with bootstrap uncertainty. Most points fall near the dashed line, which means the model predictions are generally close to the observed values. However, there is still some spread around the line, showing that the model does not perfectly predict all observations. The bootstrap results show that there is some uncertainty in the model’s predictions.